---
title: "Fetch WQP and EDR data end to end"
description: "Build a public catalog, preview its plan, and retrieve bounded provider data."
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Fetch WQP and EDR data end to end}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

This guide exercises the complete public path available today:

```text
AOI -> catalog -> fetch plan -> provider response -> typed result
```

The examples are deliberately narrow. Each selected distribution gets at most
one physical request, the scheduler is sequential, and no next page is
followed. Install the two optional provider adapters before running both
examples:

```{r install-adapters}
install.packages(c("dataRetrieval", "edr4r"))
library(geoconnexr)
```

## Fetch WQP data from a Geoconnex PID

This example starts with a real Water Quality Portal monitoring-location PID.
Supplying `site_uri` avoids a spatial graph query and retrieves that exact
JSON-LD profile. The resulting catalog records that its membership in the
Virginia AOI was not rechecked.

```{r wqp-catalog}
wqp_pid <- "https://geoconnex.us/iow/wqp/21VASWCB-WMPO001"

wqp_catalog <- gx_catalog(
  gx_aoi("VA"),
  site_uri = wqp_pid,
  max_sites = 1L
)

wqp_catalog$sites[c("site_uri", "name")]
wqp_catalog$datasets[
  wqp_catalog$datasets$fetchable,
  c("variable_name", "distribution_url", "handler_id")
]
wqp_catalog$problems[c("code", "severity", "message")]
```

Use exact UTC bounds. The end date below is December 30 because this profile's
published temporal coverage ends at the start of December 31; the planner does
not silently expand or reinterpret that boundary.

```{r wqp-fetch}
wqp_time <- as.POSIXct(
  c("2017-01-01 00:00:00", "2017-12-30 23:59:59"),
  tz = "UTC"
)

wqp_plan <- gx_fetch_plan(
  wqp_catalog,
  time = wqp_time,
  max_datasets = 1L,
  max_bytes = 1024^2
)

# Verify selection without probing packages, DNS, or providers.
gx_fetch(wqp_plan, dry_run = TRUE)$status

wqp_fetched <- gx_fetch(wqp_plan)
wqp_fetched$status

wqp_data <- wqp_fetched$results$data[[1L]]
wqp_data
```

On July 22, 2026, that request returned one row and 78 columns from 2,403
response bytes. Those dimensions are an observation, not a permanent provider
guarantee. Check `wqp_fetched$status` every time before using the payload.

## Fetch EDR CoverageJSON

The EDR execution path is live, but the tested Geoconnex SNOTEL PID currently
publishes variables without distribution URLs. Until a tested PID advertises
the reviewed EDR endpoint, this package includes a small caller-supplied
JSON-LD profile for the pygeoapi demonstration service. `gx_catalog()` records
that no PID profile request was made and does not claim the metadata came from
the provider.

```{r edr-catalog}
edr_uri <- "https://example.org/geoconnexr/icoads-sst"
profile_path <- system.file(
  "extdata", "icoads-sst-profile.json",
  package = "geoconnexr"
)
edr_profile <- jsonlite::fromJSON(profile_path, simplifyVector = FALSE)

edr_catalog <- gx_catalog(
  gx_aoi("VA"),
  site_uri = edr_uri,
  profiles = stats::setNames(list(edr_profile), edr_uri),
  max_sites = 1L
)

edr_catalog$datasets[
  , c("variable_name", "distribution_url", "handler_id", "fetchable")
]
edr_catalog$problems[c("code", "severity", "message")]
```

The profile binds a CRS84 point, one `SST` parameter, and one exact UTC
interval. Planning re-derives those facts before authorizing the request.

```{r edr-fetch}
edr_time <- as.POSIXct(
  c("2000-01-16 06:00:00", "2000-02-16 06:00:00"),
  tz = "UTC"
)

edr_plan <- gx_fetch_plan(
  edr_catalog,
  time = edr_time,
  max_datasets = 1L,
  max_bytes = 1024^2
)

gx_fetch(edr_plan, dry_run = TRUE)$status

edr_fetched <- gx_fetch(edr_plan)
edr_fetched$status

edr_data <- edr_fetched$results$data[[1L]]
edr_data
```

On July 22, 2026, that request returned two rows and nine columns from 1,908
CoverageJSON bytes. The strict package parser and
`edr4r::covjson_to_tibble()` agreed on the normalized result.

## Automatic AOI discovery

Omit `site_uri` to ask `gx_catalog()` to resolve identifier geometry and issue
one bounded `sites_in_aoi` graph query:

```{r automatic-discovery}
catalog <- gx_catalog(gx_aoi("02070010"), max_sites = 25L)
catalog$metadata$completeness
catalog$problems
```

This is implemented, but current live polygon queries can time out upstream.
Geoconnex documents graph query functionality as active development. For a
known site, the explicit-PID form is currently the reliable fallback; it is
never presented as proof that the site lies inside the AOI.

## Read the result contract

`gx_fetch()` isolates failures, so a returned object is not equivalent to
"everything succeeded." Use both tables:

- `$status` has exactly one terminal row per planned distribution, including
  attempts, bytes, and typed error codes.
- `$results` has one row per success. The handler-native payload is in the
  `data` list-column.
- `$provenance` retains the bounded request and validation evidence.

The current boundary supports direct CSV, WQP Result, EDR position, current
USGS continuous and daily data, and OGC API Features. It is single-page and
sequential by contract. Harmonization is the next roadmap milestone.
