---
title: "Get started with geoconnexr"
description: "Install the package and complete its core public workflows."
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get started with geoconnexr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

`geoconnexr` helps R users navigate connected water data without discarding
identifier identity or the evidence needed to interpret a result. This guide
uses only the package's current public functions. Fetch planning and execution
are available for the frozen supported subset, and `gx_catalog()` now connects
an AOI or explicit PID to them. Harmonization and snapshot composition remain
under active development.

## Install

Install the development version from GitHub:

```{r install}
install.packages("pak")
pak::pak("ksonda/geoconnexr")
```

Then attach the package:

```{r library}
library(geoconnexr)
```

## Start offline with an area of interest

Hydrologic identifiers stay as character values, so significant leading zeroes
are preserved. Creating a HUC recipe is deterministic and makes no network
request:

```{r aoi-huc}
aoi <- gx_aoi("02070010")
aoi$recipe
```

Custom polygon recipes are offline too. Supply exactly one valid `POLYGON` or
`MULTIPOLYGON` with an explicit coordinate reference system:

```{r aoi-polygon}
geometry <- sf::st_as_sfc(
  "POLYGON ((-77.2 38.8, -77.0 38.8, -77.0 39.0, -77.2 39.0, -77.2 38.8))",
  crs = "OGC:CRS84"
)

custom_aoi <- gx_aoi(geometry, type = "sf")
custom_aoi$recipe$aoi[c("crs", "wkb_sha256")]
```

## Resolve a persistent identifier

`gx_resolve()` follows a Geoconnex PID through the package's bounded request
layer and retains both the original PID and the resolved landing location:

```{r resolve}
pid <- "https://geoconnex.us/ref/gages/1000001"
resolution <- gx_resolve(pid)

resolution[c("pid_uri", "landing_url", "problem_code")]
```

A landing URL is not substituted for the PID as the resource's identity.

## Retrieve and parse a Geoconnex profile

Retrieve bounded JSON-LD, then parse the monitoring-location and dataset views:

```{r profiles}
document <- gx_jsonld(pid)
location <- gx_parse_location(document)
datasets <- gx_parse_datasets(document)

location
datasets
attr(location, "diagnostics")
```

Diagnostics are part of the result contract. Inspect them instead of assuming
that an empty or partial table means the upstream service had no data.

## Explore reference features

Discover the available collections and their queryable fields before issuing a
bounded feature request:

```{r reference}
gx_ref_collections()
gx_ref_queryables("hu12")

mainstem <- gx_ref_features(
  "mainstems",
  query = list(id = "29559"),
  limit = 2L
)

attr(mainstem, "gx_reference")[c("truncated", "stop_reason", "requests")]
```

Use `gx_ref_feature()` when you already know the collection and feature ID:

```{r reference-one}
gage <- gx_ref_feature("gages", "1000001")
gage
```

## Crosswalk a provider gage identifier

Provider identifiers are opaque character values. They are never coerced to
numbers:

```{r crosswalk}
gages <- gx_gage_to_pid(c("USGS-08332622", "USGS-00000000"))

gages[c("requested_provider_id", "status", "gage_uri", "mainstem_uri", "comid")]
attr(gages, "gx_crosswalk")[c("complete", "requests", "diagnostics")]
```

## Render a reviewed query template

Inspect the bundled typed templates and render one with validated parameters:

```{r query}
gx_templates()

query <- gx_render_query(
  "sites_on_mainstem",
  list(
    mainstem_uri = "https://geoconnex.us/ref/mainstems/1622734",
    limit = 100,
    offset = 0
  )
)

query
```

Rendering is local. It does not execute the query.

## Plan and fetch the supported subset

Build a validated catalog from a known PID, then build the bounded plan before
performing provider work. Explicit PIDs skip automatic graph discovery, and
the catalog records that AOI membership was not independently rechecked.

```{r fetch}
wqp_catalog <- gx_catalog(
  gx_aoi("VA"),
  site_uri = "https://geoconnex.us/iow/wqp/21VASWCB-WMPO001",
  max_sites = 1L
)

plan <- gx_fetch_plan(
  wqp_catalog,
  time = as.POSIXct(c(
    "2017-01-01 00:00:00", "2017-12-30 23:59:59"
  ), tz = "UTC"),
  max_datasets = 1L,
  max_bytes = 1024^2
)

preview <- gx_fetch(plan, dry_run = TRUE)
preview$status

fetched <- gx_fetch(plan)
fetched$status
fetched$results[c("handler_id", "payload_class", "row_count")]
fetched$results$data[[1L]]
```

The frozen M7 contract is sequential and single-page. It supports direct CSV,
WQP Result, EDR position, current USGS continuous and daily data, and OGC API
Features. Inspect every status row: one failed or unsupported distribution does
not abort later work. `fetched$results$data` retains handler-native tables or
`sf`; `fetched$provenance` retains the validated bounded execution evidence.
See [Fetch WQP and EDR data end to end](end-to-end-fetch.html) for both live
provider examples and the caller-supplied EDR profile boundary.

## Where to go next

- Read [Safety and reproducibility](safety-and-reproducibility.html) before
  building operational workflows.
- Browse the [function reference](../reference/index.html) for limits, return
  values, and error behavior.
- Follow the [build roadmap](https://github.com/ksonda/geoconnexr/blob/main/geoconnexr-spec-v0.2.md)
  for the distinction between available APIs and planned capabilities.
